---
output: rmarkdown::github_document
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/goodekat/redres/branch/master/graph/badge.svg)](https://codecov.io/gh/goodekat/redres?branch=master)
<!-- badges: end -->
  
# redres

`redres` is an R package developed to help with diagnosing linear mixed models fit using the function `lmer` from the lme4 package. This package both computes residuals types and creates plots that are not available from lme4.

## Installation

`redres` can be installed from the GitHub repository using the devtools package.

```{r eval = FALSE}
# Installs redres
devtools::install_github("goodekat/redres")
```

Once installed, the package can be loaded as usual.

```{r}
# Loads the library
library(redres)
```

## Functions 

The package contains the following functions. 

#### redres

`redres` computes residuals given an `lmer` model and a specified residual type. The available residuals types are listed below.

- `"genres"`: generalized residuals
- `"pearson_cond"`: Pearson conditional residuals
- `"pearson_mar"`: Pearson marginal residuals
- `"raw_cond"`: raw conditional residuals (default)
- `"raw_mar"`: raw marginal residuals
- `"std_cond"`: studentized conditional residuals
- `"std_mar"`: studentized marginal residuals

```{r message = FALSE}
# fits an lmer model
library(lme4)
m <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)

# computes the default residuals (raw conditional)
rc_resids <- redres(m)

# computes the Pearson marginal residuals
pm_resids <- redres(m, type = "pearson_mar")

# computes the studentized conditional residuals
sc_resids <- redres(m, type = "std_cond")

# puts the residuals in a data frame and prints the first six rows
resids <- data.frame(rc_resids, pm_resids, sc_resids)
head(resids)
```

#### plot_resid

`plot_resid` creates a plot (using `ggplot2`) of the residuals versus the fitted values given a model and a specified residual type. All residuals types listed for `redres` work with `plot_resid` except `genres` since this type of plot is not meaningful for generalized residuals.

```{r}
# creates a plot of the conditional studentized residuals versus the fitted values
plot_redres(m, type = "std_cond")
```

#### plot_genres

`plot_genres` (Kellie - want to fill in this section)

```{r}
# creates a generalized residual quantile plot
plot_genres(m)
```


#### plot_raneff

`plot_raneff`... (Kellie - want to fill in this section as well)

```{r}
# creates a normal quantile plot of the random effects
plot_raneff(m)
```

#### (shiny app function)

`(shiny app function)`...to be filled in in once available 

## Learn More

To learn more about `redres`, read through the [vignette](https://goodekat.github.io/redres/articles/redres-vignette.html) for `redres` which contains:

- details on how the residuals are computed 
- explanations of how to use the different residual types 
- interpretations of the plots
- an example using `redres` to assess the assumptions of a linear mixed model

Additionally, more information is available at the package [website](https://goodekat.github.io/redres/).
