---
title: "redres: Redress Your Mixed Model Assumptions"
author: "R Package Version `r packageVersion('redres')`"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 5
)
```

# Introducing redres

## Overview

(motivation for the package on having tools for diagnosing mixed models) The Merriam-Webster dictionary defines [redress](https://www.merriam-webster.com/dictionary/redress) as "to set right".

## Package Structure

(describe the three ways to use the package)

`redres` contains various functions that provide different ways to assess model assumptions for linear mixed models fit using `lmer`. The function range from returning a vector of residuals to be used by the user as desired to opening a Shiny app that allows the user to interactively assess the model with package provided tools. A brief description of each function is listed below.

- [redres](#redres): computes and returns residuals given a model and a specified residual type
- [plot_redres](#plot_redres): creates a plot of residuals versus fitted values for a given residual type
- [plot_resqq](#plot_resqq): creates a normal quantile plot for conditional residuals
- [plot_raneff](#plot_raneff): creates a normal quantile plot for the random effects
- [redres_app](#redres_app): opens an interactive Shiny app with 

## Installation

The package can be installed from GitHub using devtools and then loaded in the normal way.

```{r eval = FALSE}
# Installs redres from GitHub
devtools::install_github("goodekat/redres")
```

```{r}
# Loads the package
library(redres)
```

## Example Data

```{r}
# Prints the first 6 rows of the paprika data
head(paprika)

# Loads the lme4 libraries
library(lme4)
m <- lmer(height ~ rep + treatment*variety + 
            (1|rep:treatment) + 
            (1|rep:treatment:variety), 
          data = paprika)
```

# Extracting Residuals

One of the ways to utilize `redres` is to extract residuals from a model fit using `lmer`. This can be done using the main function in the package named `redres` to compute residuals based on a specified type.

## Residual Types {#types}

The linear mixed effects model can be written as
  $$\textbf{Y}=\textbf{X}\boldsymbol{\beta}+\textbf{Z}\boldsymbol{\gamma}+\boldsymbol{\epsilon}$$
where

- $\textbf{Y}$ is an $n\times 1$ vector of $n$ response variable observations,
- $\textbf{X}$ is an $n\times p$ matrix of $p$ explanatory variables with $n$ observations each,
- $\boldsymbol{\beta}$ is a $p\times1$ vector of unknown fixed effects parameters,
- $\textbf{Z}$ is an $n\times q$ matrix of $q$ random effect variables with $n$ observations each,
- $\boldsymbol{\gamma}$ is a $q\times1$ vector of unknown random effects, and
- $\boldsymbol{\epsilon}$ is an $n\times1$ vector of random errors.

Under this set up,
  $$E[\textbf{Y}]=\textbf{X}\boldsymbol{\beta} \ \ \ \mbox{ and } \ \ \ 
  Var[\textbf{Y}]=\textbf{ZGZ}'+\textbf{R}=\textbf{V}.$$
  
It is assumed that
$$ 
\begin{bmatrix} \boldsymbol{\gamma} \\ \boldsymbol{\epsilon} \end{bmatrix}
\sim N 
\begin{pmatrix} 
\begin{bmatrix} \boldsymbol{0} \\ \boldsymbol{0} \end{bmatrix},
\begin{bmatrix} \textbf{G} & \boldsymbol{0} \\ \boldsymbol{0} & \textbf{R}
\end{bmatrix}
\end{pmatrix}.
$$

`redres` can compute six types of residuals. These include some types not available in `lme4`, and some of the types are computed differently in `redres` than `lme4`. These changes were made to make the choices of type names more intuitive with the computations used. The formulas for how these are computed in `redres` are listed below.

### Raw Residuals

The raw residuals are computed as the observed response values minus the predicted response values where the marginal version does not account for the random effects while the conditional version does. These are computed as follows.

\vspace{0.25cm}

- **marginal** raw residuals
  $$r^m_i = Y_i-\textbf{x}'_i\widehat{\boldsymbol{\beta}}$$
  
- **conditional** raw residuals
  $$r^c_i = Y_i-\textbf{x}'_i\boldsymbol{\widehat{\beta}}-\textbf{z}'_i\widehat{\boldsymbol{\gamma}}$$

### Pearson Residuals

The Pearson residuals are computed as the raw residuals divided by the square root of the estimated variance of the response values. Both marginal and conditional versions are available as follows.

\vspace{0.25cm}

- **marginal** Pearson residuals
  $$r^{m,Pearson}_{i} = \frac{r^m_i}{\sqrt{\widehat{Var}[Y_i]}}$$
  
- **conditional** Pearson residuals
  $$r^{c,Pearson}_{i} = \frac{r^c_i}{\sqrt{\widehat{Var}[Y_i|\boldsymbol{\gamma}]}}$$

### Studentized

The studentized residuals are computed as the raw residuals divided by the square root of the estimated variance of the raw residuals. Again, both marginal and conditional versions are available.

- **marginal** Pearson residuals
  $$r_i^{m,std}=\frac{r_i^m}{\sqrt{\widehat{Var}[r_i^m]}}$$
  
- **conditional** studentized residuals
  $$r_i^{c,std}=\frac{r_i^c}{\sqrt{\widehat{Var}[r_i^c]}}$$

Note that
  $$\widehat{Var}[\textbf{r}^m]=\widehat{\textbf{V}}-\textbf{Q} \ \ \ \mbox{ and } \ \ \ \widehat{Var}[\textbf{r}^c]=\textbf{K}\left(\widehat{\textbf{V}}-\textbf{Q}\right)\textbf{K}'.$$
The values of $\textbf{Q}$ and $\textbf{K}$ are defined as follows by by Gregoire, Schabenberger, and Barrett (1995).

$$\textbf{Q}=\textbf{X}(\textbf{X}'\widehat{\textbf{V}}^{-1}\textbf{X})^{-}\textbf{X}' \ \ \ \mbox{ and } \ \ \ \textbf{K}=\textbf{I}-\textbf{Z}\widehat{\textbf{G}}\textbf{Z}'\widehat{\textbf{V}}^{-1}$$

## Usage

The function `redres` can compute the residuals types defined in the previous section for models fit using `lmer`.

**Inputs**

- `model`: A model fit using `lmer` for which the residuals will be computed.
- `type`:	A character string identifying the type of residual that will be computed. By default, the raw conditional residuals are returned. The following are the options available for type.
      - `"pearson_cond"`: Pearson conditional residuals
      - `"pearson_mar"`: Pearson marginal residuals
      - `"raw_cond"`: raw conditional residuals (default)
      - `"raw_mar"`: raw marginal residuals
      - `"std_cond"`: studentized conditional residuals
      - `"std_mar"`: studentized marginal residuals  
      
See the section on [residual types](#types) for details on how the residuals are computed and their purpose.

**Output**

- `redres` returns a vector of residuals according to the type specified in the order that the data observations are input into `lmer`.

**Functionality** 

The code below demonstrates the use of `redres` to compute several types of residuals from the model `m`. These residuals are put into a dataframe and joined with the `sleepstudy` data.

```{r}
# Computes the default residuals (raw conditional)
raw_cond <- redres(m)

# Computes the Pearson marginal residuals
pearson_mar <- redres(m, type = "pearson_mar")

# Computes the studentized conditional residuals
std_cond <- redres(m, type = "std_cond")

# Joins the residuals to the paprika data
paprika_plus <- cbind(paprika, raw_cond, pearson_mar, std_cond)

# Prints the head of the dataframe
head(paprika_plus)
```

The user can then use these residuals as desired. For example, the user could make their own plots or perform tests on the residuals. The code below creates histograms of the three residual types.

```{r message = FALSE}
# Loads helpful libraries 
library(dplyr)
library(tidyr)
library(ggplot2)

# Creates histograms of the residual types
paprika_plus %>%
  gather(key = "type", value = "residual", 6:8) %>%
  ggplot(aes(x = residual)) + 
  geom_histogram(bins = 30) + 
  facet_grid(. ~ type, scales = "free") + 
  theme_bw()
```

# Plotting Residuals

## Residual Plots 

## Quantile plots

Quantile plots are used to verify visually if data follows are particular distribution. Data points are plotted along the quantiles of the assumed distributed. The data is expected to follow an approximately straight line, at least along the middle quantiles. The points are assessed for any extreme curvation that would indicate a departure from the assumed distribution. Typically curvature at the extreme ends of the quantiles (around 0 and 1) are ignored as data is sparse here and distributions behave more erratically at the boundaries of the parameter space. We have added confidence bands to guide the user.  

The function `plot_ranef` plots each random effect vector along the normal quantiles. From the assumptions of the linear mixed model, each random effect specified - along with the error term - is assumed to follow a normal distribution. Therefore, these plots can be used to assess if this assumption is met. Note that the number of plots generated by this function will vary for each model, with the number of plots being the number of random effects.

## Usage

### plot_redres {#plot_redres}

The function `plot_redres` is included in the package to allow for the easy creation of a residual plot using `ggplot2` for a linear mixed model fit using `lmer`. `plot_redres` allows the user to specify any residual type to be plotted on the y-axis including those not provided by `lme4`.

**Inputs**

- `model`: A model fit using `lmer` for which the residuals will be computed.
- `type`:	A character string identifying the type of residual that will be computed. The same options available with [`redres`](#redres) are available for `plot_redres`. By default, the conditional raw residuals are plotted.

**Output**

- A `ggplot2` object of a scatterplot of the model residuals of the type specified versus the fitted values is created.

**Functionality**

The code below shows how to create the most basic plot using `plot_redres`. By inputting the model `m`, a plot of the conditional raw residuals versus the fitted values is returned. 

```{r}
# Creates the default residual plot
plot_redres(m)
```

The type of residuals can be changed using the `type` option as shown in the code below.

```{r}
# Creates the residual plot with studentized marginal residuals
plot_redres(m, type = "std_mar")
```

All plots created using the `redres` package are formatted to use `theme_bw` from `ggplot2`. However, since `plot_redres` returns a ggplot object, it is possible to use functions provided by `ggplot2` adjust the formatting of the plot.

```{r}
# Applies ggplot2 formatting functions to the output from plot_resid
library(ggplot2)
plot_redres(m, type = "pearson_cond") +
  theme_classic() +
  labs(title = "Residual Plot")
```

### plot_resqq {#plot_resqq}

### plot_ranef {#plot_ranef}

# Interacting with Residuals

## Shiny App {#redres_app}

The function `redres_app` includes a shiny app that helps demonstrate how the plotting functions `plot_redres`, `plot_resqq` and `plot_ranef` can be used. It can also be used to visually comparing two linear mixed models and conducting model selection through pairwise comparison.

**Inputs**

- `model`: A model (or two models wrapped in a list) fit using `lmer`.

**Output**

- A `shiny` app.

**Functionality** 

The code below shows how to create a `shiny` app. By inputting the model `m` (could be a list of two models), a `shiny` app which has two main tabs, one includes the residual plot with a choice of all [residual types](#types) and a choice of different x-axis variables and another one includes the normal quantile plots is returned.

## Usage





